The Galactic Midplane Is Not a Plane: Implications for Dynamical Analysis with {\em Gaia} Data and Beyond

### 2019-04-24 paper/ms.tex
Remove neighborhood stuff in paper

In Section~\ref{sec:discussion} we discuss the implications of midplane
variations, and the resulting systematic uncertainty in the vertical action,
for action-space analyses beyond the dynamical solar neighborhood. As an
illustration, in Section \ref{sssec:neighborhood} we use these estimates to
consider the feasibility of one science goal that relies on the invariance of
actions---the reconstruction of open clusters across large distances using
``dynamical tagging;'' that is, the association of stars to a birth cluster
based on their similar actions. In light of our findings, we considered the
orbits of the nine open clusters within $250\,\pc$ as reported by the {\em
Gaia} collaboration \citep{2018AA...616A..10G}. In
Table~\ref{tab:real_clusters}, we report the distance to each cluster, the
three actions, and the maximum vertical extent of their orbits
($z_{\text{max}}$). In Section~\ref{sssec:reconstruction} we estimate and
discuss the size of the dynamical solar neighborhood for each cluster based on
its mass, to show that action-space tagging is likely still a viable method if
contamination from foreground or background stars is unimportant. We summarize
and conclude in Section~\ref{sec:conclusion}.

@@

\subsection{The Size of the Solar Neighborhood} \label{sssec:neighborhood}
Given a local coordinate system, the amplitude and characteristic length scale
of the density fluctuations in the Galactic disk will determine how far this
coordinate system can be extended before introducing relevant systematic
errors in the computation of actions for thin-disk stars. We consider a $10\%$
error to be a criterion for the region around the Sun where the analysis of
the action-space distribution of stars on quasi-circular orbits, and therefore
the epicyclic approximation, can be expected to be valid. This leads to a
somewhat dynamically motivated definition of the ``solar neighborhood.'' Since
the motivation may be different for defining the solar neighborhood in, e.g.,
studies of the chemical distribution of the Galaxy, we propose the phrase
``dynamical solar neighborhood.'' We also point out that our use of a $10\%$
error is arbitrary.

To illustrate this concept, we first compute the absolute range of midplane
values for a given $\Delta \phi$ segment of the midplane. With the observer's
azimuth set to zero, we consider a region extending from $-\Delta \phi/2$ to
$\Delta\phi/2$. To translate this angular scale into a physical scale, we also
compute the chord length $\lambda$ for the chord extending from the observer
to $\pm \Delta\phi/2$ (see Figure~\ref{fig:fig_to_explain}). We repeat this
for each of $50$ bins in azimuth. Each bin is plotted in
Figure~\ref{fig:range_deltaphi} as a gray line, with the mean value in solid
blue and the $1\,\sigma$ dispersion as the dashed blue lines. We choose to
only consider the Solar circle to make the point that the midplane variation
as a function of azimuth (and not just Galactic radius) is important. However,
we note that because we use the range of midplane values, including regions of
each galaxy outside the Solar circle can only act to increase this range, and
therefore reduce the size of the dynamical solar neighborhood.

\begin{figure*}
\plotone{fig/fig_to_explain.pdf}
\caption{A cartoon explanation of Figure~\ref{fig:range_deltaphi}. An observer
is placed at the blue x in the plane of a backgrround galaxy (lime). For a
given $\Delta \phi$ (blue) centered on the observer, we record the range of
midplane values for this section of the Solar circle (orange). We plot the
value of the range against the value of $\Delta \phi$. We are also able to
convert the value of $\Delta \phi$ into a chord length (pink), which we plot
as a secondary, upper $x$-axis. We repeat the procedure for each initial
$\phi$ (gray lines), and also compute the mean and $\pm1\,\sigma$ values (dark
blue). We only consider the Solar circle to explore the impact of azimuthal
midplane variations, but similar effects are expected for radial midplane
variations.}
\label{fig:fig_to_explain}
\end{figure*}

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/range_dphi.pdf}
\end{center}
\caption{The range of midplane heights encountered as a function of angular
width. At each angle $\phi$ from Figure~\ref{fig:midplane} we consider an
angular width of $\Delta \phi$ centered on $\phi$ and report the range of
midplane heights within that width. We repeat the procedure for each $\phi$
and plot the result as translucent gray lines. We also plot the mean range as
a solid blue line and the $\pm1\sigma$ lines as dashed blue lines. The upper
$x$-axis shows the chord length from the position $\phi$ to $\pm\Delta\phi/2$.
The horizontal \thincolor{} and \thickcolor{} lines shows the midplane
variation necessary to induce a $10\%$ error in $J_z$ for the thin-disk and
thick-disk orbits, respectively. A cartoon explanation of this Figure is given
in Figure~\ref{fig:fig_to_explain}. We only consider the Solar circle to
explore the impact of azimuthal midplane variations, but similar effects are
expected for radial midplane variations.}
\label{fig:range_deltaphi}
\end{figure*}

From Figure~\ref{fig:range_deltaphi} it is clear that the range of midplane
values spanned in a given $\Delta \phi$ quickly saturates at $\Delta \phi
\gtrsim \pi/2$ radians (i.e. one quarter of the way around the Galactic disk
from the Sun). At far smaller $\Delta \phi$ it reaches the level of deviation
required for a $10\%$ systematic error in $J_z$ for our fiducial
thin-disk-like orbit (\thincolor{} horizontal line in Figure
\ref{fig:range_deltaphi}). For the three simulations considered, the median
range exceeds this value at $\Delta \phi/\pi > 0.0073, 0.0096,
0.018\,\text{rad}$, with corresponding physical scales of $\lambda \sim 93.9,
123, 233\,\pc$ for \mi{}, \mf{}, and \mm{}, respectively. These can be
considered estimates of the range of validity for an extension of the locally
determined axisymmetric coordinate system, at least when it comes to the study
of stars on quasi-circular orbits.

The situation is more forgiving for the study of thick-disk-like orbits, where
a $10\%$ systematic error in $J_z$ requires a larger midplane
deviation (\thickcolor{} horizontal line in Figure~\ref{fig:range_deltaphi}).
In fact, the range of midplane variation in \mi{} just barely saturates to
cause such an error in $J_z$, with the midplane variation in \mf{} and \mm{}
never reaching such a value. For stars on halo-like orbits, extending the
local coordinate system globally is not likely to cause serious problems,
since the midplane deviations are relatively small compared to those needed to
produce $10\%$ systematic errors in the actions (\halocolor{}
horizontal line in Figure \ref{fig:range_deltaphi}). However, we caution that
the tendency of these midplane deviations is to \emph{inflate} the value of
$J_z$ whenever the midplane deviation is $\gtrsim z_{\text{max}}$, scattering
them into the region of action space normally associated with different
stellar populations and possibly contributing to confusion there (see
Figure~\ref{fig:Jz_hist} and discussion therein). This inflation is extremely
non-Gaussian (as the induced error is for all orbits, e.g.
Figure~\ref{fig:Jz_hist}), and is particularly troubling for the case of thin
disk stars, where the necessary midplane deviation is
$\sim5\textup{--}10\,\pc$.

@@

With a characteristic midplane offset in hand, we use the relation in
Figure~\ref{fig:range_deltaphi}, which shows the typical midplane offset at
each $\Delta \phi$, to convert a given midplane offset into a typical
$\Delta\phi$. As in this Figure we assume the cluster is at the Solar circle
to obtain a physical scale, which we refer to as the ``neighborhood radius''
or $R_{\n}$. If the allowable midplane discrepancy is greater than the maximum
excursions recorded in the simulated galaxy, we simply report the maximum
chord length assuming an orbit on the Solar circle ($2\times8.2\,\kpc$). If an
observer is attempting to reconstruct a cluster with a given mass on a given
orbit, $R_{\n}$ gives the characteristic search distance between a field star
and the cluster within which an observer may not need to consider differences
in their respective local midplanes when computing actions. The result is
shown in Figure~\ref{fig:Rn_mc}. The solid lines indicate the neighborhood
values assuming the mean relation between midplane offset and $m_c$, and the
dashed lines show the $+1\,\sigma$ and $-1\,\sigma$ relations based on the 50
fiducial Solar positions in each simulation (as discussed in Figure
\ref{fig:range_deltaphi}). That this range is relatively wide illustrates the
strong $\phi$-dependence of the midplane offsets.

@@

Figure~\ref{fig:Rn_mc} shows that azimuthal midplane fluctuations will limit
the volume over which actions can be used to match stars to open clusters on
thin-disk-like orbits (\thincolor), depending on their mass. For
thick-disk-like orbits (\thickcolor), the tolerances are much more forgiving:
stars' actions can be compared over the entire Solar circle for nearly all
cluster masses, except at the very low-mass end in \mi . The neighborhood
around $\sim100\Msun$ clusters can be as small as $\sim x\,\pc$ but up to
$\sim y\,\kpc$ for \mi{} and \mf{}, though the situation is marginally better
for \mm{} with values ranging from $\sim a\,\kpc$ to $\sim b\,\kpc$. The
neighborhood values steadily increase with $m_c$, topping out at $\sim c\,\kpc$
for $\sim10^4\,\Msun$ clusters in
\mi{} and \mf{}. In \mm{}, the neighborhood size reaches the maximum allowed of
$16.4\,\kpc$ for clusters of mass $\sim m\,\Msun$. This figure also
underlines that the degree to which such an analysis will succeed depends
crucially on the expected degree of midplane fluctuations in the real Galaxy.
If the Milky Way is more like \mi , the range of action-space analysis
techniques like this is far more limited than if it more closely resembles
\mm{}.

We also compute $R_{\n}$ for each of the orbits of the real open clusters in
Table~\ref{tab:real_clusters}, using the midplane variations from
\mi{} as a conservative choice, since it has the largest degree of
midplane variation (Figure~\ref{fig:midplane}). We report the neighborhood
values for the mass of each cluster in the neighborhood column of
Table~\ref{tab:real_clusters}. Except for alphaPer, we find that the
neighborhood around each cluster is typically $\gtrsim x\,\pc$. This analysis
assumes that the MW resembles \mi{}, which is not necessarily true. However,
this step of our analysis is meant to be more illustrative than quantitative:
this calculation should actually be performed using real measurements of
variations in the MW's stellar midplane. Such measurements should be possible
in the near future with {\em Gaia}, with quality distance estimates the
crucial limiting factor, along with careful selction function modelling.

\begin{figure*}
\begin{center}
\includegraphics[width=\textwidth]{fig/Rn_vs_mc.pdf}
\end{center}
\caption{The dynamical neighborhood (discussed in
Section~\ref{sssec:neighborhood}) around an open cluster of mass $m_c$ for the
thin-disk, and thick-disk orbits (\thincolor{} and \thickcolor{},
respectively; see Table~\ref{tab:orbits}) and for each of the FIRE galaxies
considered here ---
\mi{}, \mf{}, and \mm{} ({\em left}, {\em center}, and {\em right} panels,
respectively). Given the $z$ offset for a cluster of mass $m_c$ computed in
Figure~\ref{fig:cluster_offset}, we convert this to a chord length using the
result from Figure~\ref{fig:range_deltaphi}. We interpret this chord length as
the ``neighborhood'' around the cluster, i.e. the distance from the cluster
one could go before the $J_z$ error induced by our midplane effect is
comparable to the intrinsic action dispersion of the cluster. When the $z$
offset is larger than the maximum range in Figure~\ref{fig:range_deltaphi}, we
report the maximum chord length ($2\times8.2\,\kpc$). We aso report the same
procedure but using the $\pm1\sigma$ lines from
Figure~\ref{fig:range_deltaphi}, which we plot here as dashed lines. For the
thick orbit (\thickcolor), the neighborhood is quite large and thus the effect
is negligible. However, for the thin orbit (\thincolor), which is closest to
the orbits of nearby open clusters (see Table~\ref{tab:real_clusters}), the
neighborhood is only a few hundred $\pc$ for low-mass clusters.}
\label{fig:Rn_mc}
\end{figure*}

>>>>

In Section~\ref{sec:discussion} we discuss the implications of midplane
variations, and the resulting systematic uncertainty in the vertical action,
for action-space analyses beyond the dynamical solar neighborhood. In light of our findings, we considered the
orbits of the nine open clusters within $250\,\pc$ as reported by the {\em
Gaia} collaboration \citep{2018AA...616A..10G}. In
Table~\ref{tab:real_clusters}, we report the distance to each cluster, the
three actions, and the maximum vertical extent of their orbits
($z_{\text{max}}$). In Section~\ref{sssec:reconstruction} we estimate and
discuss the size of the dynamical solar neighborhood for each cluster based on
its mass, to show that action-space tagging is likely still a viable method if
contamination from foreground or background stars is unimportant. We summarize
and conclude in Section~\ref{sec:conclusion}.

@@

@@

@@
